### 2021-12-06 manuscript.tex
Add ilustrative images for the session on Wikidata Bib [ci skip]
This build is based on
https://github.com/lubianat/quali_phd/commit/856180b09ff9ec0998a2a1d9e24bc3364473fbc9.

This commit was created by the following CI build and job:
https://github.com/lubianat/quali_phd/commit/856180b09ff9ec0998a2a1d9e24bc3364473fbc9/checks
https://github.com/lubianat/quali_phd/actions/runs/1546263356

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/d4734c3110210b188b6c531653e9e446bd8e91b4/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/d4734c3110210b188b6c531653e9e446bd8e91b4}{lubianat/quali\_phd@d4734c3}
on December 3, 2021.

@@

\begin{itemize}
\item
  \textbf{image note-taking example}
\item
  \emph{Talk about Author Disambiguator and other contributions}
\end{itemize}

@@

\begin{itemize}
\item
  Wikidata Bib generates real-time, public visualizations of my literature progress
\item
  wbib package: https://github.com/lubianat/wbib
\item
  \textbf{Figures for queries}
\item
  \emph{Wikidata Bib reading status}:

@@

\begin{figure}
\hypertarget{fig:biocuration_of_cells}{%
\centering
\includegraphics[width=0.85\textwidth,height=\textheight]{images/extracting_cell_types.svg}
\caption{Wikidata Bib was coupled with a biocuration framework for cell types}\label{fig:biocuration_of_cells}
}
\end{figure}

@@

\hypertarget{preliminary-results-1}{%
\section{Preliminary Results}\label{preliminary-results-1}}

\hypertarget{fcoex-updates}{%
\subsection{Fcoex updates}\label{fcoex-updates}}

\hypertarget{the-concept-of-cell-type}{%
\subsection{The concept of cell type}\label{the-concept-of-cell-type}}

\begin{itemize}
\tightlist
\item
  Describe background
\item
  Cell types, cell states and cell classes
\item
  Levels of cell type information: archetype, senso stritu cell type, infratype and technotype.
\item
  Infratypes and technotypes as theoretical innovations
\item
  Current usage mixes archetypes and species-specific cell types
\item
  Annotation of HCA articles for grasping the use of different concepts in the context of HCA
\end{itemize}

\hypertarget{next-steps}{%
\subsubsection{Next steps}\label{next-steps}}

\begin{itemize}
\tightlist
\item
  Improve formalization of cell types in connection with the biomedical semantics community
\end{itemize}

\hypertarget{hca}{%
\subsection{HCA}\label{hca}}

\begin{itemize}
\tightlist
\item
  ``Sky dive'' approach: hand annotation of all abstracts and the core Human Cell Atlas paper
\item
  Benefits of using a single ontology that anyone can edit (new terms and speed of science)
\item
  Figure: The different concepts in use by the HCA paper
\item
  Figure: The different concepts in use by the different HCA papers
\item
  Discussion
\item
  Information by HCA and related efforts is already targeted by biocurators. PanglaoDB is one of
  these resources etc etc
\end{itemize}

\hypertarget{next-steps-1}{%
\subsubsection{Next steps}\label{next-steps-1}}

\begin{itemize}
\tightlist
\item
  Mature the annotation system into a curation tool (based on ANN, perhaps reuse figure)
\item
  Explore the use of SciSpacy and natural language processing for making it easier
\end{itemize}

\hypertarget{cell-ontology}{%
\subsection{Cell Ontology}\label{cell-ontology}}

\hypertarget{panglaodb-integration-to-wikidata-1}{%
\subsection{PanglaoDB integration to Wikidata}\label{panglaodb-integration-to-wikidata-1}}

\begin{itemize}
\tightlist
\item
  The architecture of marker information on Wikidata
\item
  Integration of information to the larger scope --\textgreater{} live updates by everyone
\item
  Overview of the stats
\end{itemize}

\hypertarget{preliminary-results-2}{%
\section{Preliminary Results}\label{preliminary-results-2}}

\hypertarget{fcoex-updates-1}{%
\subsection{Fcoex updates}\label{fcoex-updates-1}}

\hypertarget{the-concept-of-cell-type-1}{%
\subsection{The concept of cell type}\label{the-concept-of-cell-type-1}}

\begin{itemize}
\tightlist
\item
  Describe background
\item
  Cell types, cell states and cell classes
\item
  Levels of cell type information: archetype, senso stritu cell type, infratype and technotype.
\item
  Infratypes and technotypes as theoretical innovations
\item
  Current usage mixes archetypes and species-specific cell types
\item
  Annotation of HCA articles for grasping the use of different concepts in the context of HCA
\end{itemize}

\hypertarget{next-steps-2}{%
\subsubsection{Next steps}\label{next-steps-2}}

\begin{itemize}
\tightlist
\item
  Improve formalization of cell types in connection with the biomedical semantics community
\end{itemize}

\hypertarget{hca-1}{%
\subsection{HCA}\label{hca-1}}

\begin{itemize}
\tightlist
\item
  ``Sky dive'' approach: hand annotation of all abstracts and the core Human Cell Atlas paper
\item
  Benefits of using a single ontology that anyone can edit (new terms and speed of science)
\item
  Figure: The different concepts in use by the HCA paper
\item
  Figure: The different concepts in use by the different HCA papers
\item
  Discussion
\item
  Information by HCA and related efforts is already targeted by biocurators. PanglaoDB is one of
  these resources etc etc
\end{itemize}

\hypertarget{next-steps-3}{%
\subsubsection{Next steps}\label{next-steps-3}}

\begin{itemize}
\tightlist
\item
  Mature the annotation system into a curation tool (based on ANN, perhaps reuse figure)
\item
  Explore the use of SciSpacy and natural language processing for making it easier
\end{itemize}

\hypertarget{cell-ontology-1}{%
\subsection{Cell Ontology}\label{cell-ontology-1}}

\hypertarget{panglaodb-integration-to-wikidata-2}{%
\subsection{PanglaoDB integration to Wikidata}\label{panglaodb-integration-to-wikidata-2}}

\begin{itemize}
\tightlist
\item
  The architecture of marker information on Wikidata
\item
  Integration of information to the larger scope --\textgreater{} live updates by everyone
\item
  Overview of the stats
\end{itemize}

\hypertarget{additional-works}{%
\section{Additional Works}\label{additional-works}}

\hypertarget{fcoex}{%
\subsection{fcoex}\label{fcoex}}

\hypertarget{cell-ontology---minimal-information-about-a-new-cell-type}{%
\subsection{Cell Ontology - Minimal Information About a New Cell Type}\label{cell-ontology---minimal-information-about-a-new-cell-type}}

\hypertarget{cellosaurus-and-wikidata}{%
\subsection{Cellosaurus and Wikidata}\label{cellosaurus-and-wikidata}}

\hypertarget{complex-portal-and-wikidata}{%
\subsection{Complex Portal and Wikidata}\label{complex-portal-and-wikidata}}

\hypertarget{wikiproject-elixir}{%
\subsection{WikiProject ELIXIR}\label{wikiproject-elixir}}

\hypertarget{systematic-reviews-and-publishing-of-intermediary-tables}{%
\subsection{Systematic Reviews and publishing of intermediary tables}\label{systematic-reviews-and-publishing-of-intermediary-tables}}

\hypertarget{academic-curriculum}{%
\section{Academic Curriculum}\label{academic-curriculum}}

\hypertarget{university-course}{%
\subsection{University course}\label{university-course}}

\hypertarget{awards-and-participation-in-events}{%
\subsection{Awards and Participation in events (?)}\label{awards-and-participation-in-events}}

\begin{itemize}
\tightlist
\item
  ISCB 2021
\item
  BioHack EU 2021
\end{itemize}

@@

\leavevmode\vadjust pre{\hypertarget{ref-xWhvkd3B}{}}%
\CSLLeftMargin{22. }
\CSLRightInline{\textbf{An estimation of the number of cells in the human body.}
\CSLBlock{Eva Bianconi, Allison Piovesan, Federica Facchin, Alina Beraudi, Raffaella Casadei, Flavia Frabetti, Lorenza Vitale, Maria Chiara Pelleri, Simone Tassani, Francesco Piva, \ldots{} Silvia Canaider} \emph{Annals of Human Biology} (2013-07-05) \url{https://www.wikidata.org/wiki/Q34037445}
\CSLBlock{DOI: \href{https://doi.org/10.3109/03014460.2013.807878}{10.3109/03014460.2013.807878}}}

@@

>>>>

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/856180b09ff9ec0998a2a1d9e24bc3364473fbc9/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/856180b09ff9ec0998a2a1d9e24bc3364473fbc9}{lubianat/quali\_phd@856180b}
on December 6, 2021.

@@

\begin{figure}
\hypertarget{fig:notetaking}{%
\centering
\includegraphics{images/note_taking_station_annotated_with_links.png}
\caption{Wikidata Bib's platform for note taking}\label{fig:notetaking}
}
\end{figure}

\begin{itemize}
\tightlist
\item
  \emph{Talk about Author Disambiguator and other contributions}
\end{itemize}

@@

\begin{itemize}
\item
  Wikidata Bib generates real-time, public visualizations of my literature progress
\item
  wbib package: https://github.com/lubianat/wbib
\end{itemize}

\begin{figure}
\hypertarget{fig:dashboard}{%
\centering
\includegraphics{images/wikidata_bib_display.png}
\caption{Wikidata Bib queries for institutions of authors and most read venues}\label{fig:dashboard}
}
\end{figure}

\begin{itemize}
\tightlist
\item
  \emph{Wikidata Bib reading status}:

@@

\begin{figure}
\hypertarget{fig:biocuration_of_cells}{%
\centering
\includegraphics[width=0.85\textwidth,height=\textheight]{images/biocuration_of_cells.png}
\caption{Wikidata Bib was coupled with a biocuration framework for cell types}\label{fig:biocuration_of_cells}
}
\end{figure}

@@

\hypertarget{final-considerations-and-next-steps}{%
\section{Final considerations and next steps}\label{final-considerations-and-next-steps}}

To sum up, this PhD research project aims at improving knowledge representation in the context of the Human Cell Atlas.
It is composed by a mixture of theoretical studies on conceptual modelling, practical contributions to knowledge organization projects, (mainly the Cell Ontology and Wikidata), explorations of the data to generate biomedical insights and the development of a technical framework for organized reading.
By approaching the object of study from a new perspective, we hope not only to make sizeable contributions, but to promote discussion and fruitful conflation of approaches.

The next years of study will be devoted to improving the projects presented here into mature, useful objects.
We hope to improve the interplay of Wikidata and Cell Ontology, developing frameworks to combine community- and expert- based curation of knowledge on cell types.
Furthermore, we plan to integrate Wikidata to current single-cell RNA-sequencing pipelines by adapting ontology-based R packages ( asOnClass {[}\protect\hyperlink{ref-sW6aNZJB}{50}{]} and ontoProc{[}\protect\hyperlink{ref-15YmDXALp}{51}{]})) to use Wikidata.
Finally, we aim at moving the Wikidata Bib system to a well documented, user-friendly mature system, testing usability with other academics and distributing it as a durable open-source project.

\hypertarget{additional-work}{%
\section{Additional Work}\label{additional-work}}

\hypertarget{collaborations-and-manuscripts}{%
\subsection{Collaborations and manuscripts}\label{collaborations-and-manuscripts}}

\hypertarget{fcoex}{%
\subsubsection{fcoex}\label{fcoex}}

During the initial course of this PhD work, we also completed the development and reportin of \emph{fcoex}, an R package for investigating cellular phenotypes using co-expression networks. {[}\protect\hyperlink{ref-MxIeSJYt}{135}{]} The software was mantained to withstand new releases of dependencies and new R version, and \emph{WAS PUBLISHED AS A PRE\_PRINT ADD HERE THE LINK}.

\hypertarget{wikidata-bots}{%
\subsubsection{Wikidata Bots}\label{wikidata-bots}}

Alongside the editing of cell-type information on Wikidata, I have joined different efforts to improve biological information on Wikidata.
I have collaborated with the ComplexPortal curators, as part of the Virtual Elixir BioHackathon 2020 (https://github.com/virtual-biohackathons/covid-19-bh20/wiki) and for the following year, to build an Wikidata Bot to integrate information on protein complexes to Wikidata. An overview of the Wikidata integration is in Figure \ref{fig:complexportal}, presented in an article published in Nucleic Acid Research (re-use of the image and legend possible under the CC-BY license of the article). {[}\protect\hyperlink{ref-CQRJ53gu}{136}{]}
\includegraphics{https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/nar/PAP/10.1093_nar_gkab991/2/gkab991fig3.jpeg?Expires=1641821957\&Signature=RK-es18S~Qh6vGQE~61i6u4prMuij8kVTbrjN6WUJLfYHOAhUhx9qQorBxROohjLLxbHvZ2YK9e7EwlI9HjVeNoGZ2PJs0Pv78Y31MdZLY8FeLYI2E4azwrqRyv9q0AH8QL3RorWZV1AhOb9bl-44Mr97Q~9MWzeTDnQQbxpCnGLG~YoG49kocD5KE~dmTSQdkXBU7kZnuGM1NPqMHo5ZDUoCRFwmTbLvd4kXoH~6CTyqx4ruQRIO-ks4Q0FUjYnxGX18gCFvoO2LMPZUPrQjTgD35RrVAkLs91Ur6BQHUuZ-ZsE3Iag3mBDu27vT9Jr93bOj0i9wz~surTtEGATgw__\&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA}

I have also collaborated with the Cellosaurus database {[}\protect\hyperlink{ref-1DguATd9G}{137}{]} to revive the CellosaurusBot {[}\protect\hyperlink{ref-lMQxhx3q}{138}{]}, responsible for updating the metadata on more than 100,000 cell lines on Wikidata. The bot code, written in Python, was completely refactored, and is run by me semi-automatically after the Cellosaurus database releases. A write-up of he integration is in progress, and is planned for release/submisson in the first semester of 2022.

\hypertarget{systematic-reviews-and-publishing-of-intermediary-tables}{%
\subsubsection{Systematic Reviews and publishing of intermediary tables}\label{systematic-reviews-and-publishing-of-intermediary-tables}}

Finally, in a collaboration with Olavo Amaral and Kleber, from the Brazilian Reproducibility Initiative {[}\protect\hyperlink{ref-F2mYjDJ0}{139}{]} I wrote a commentary on the value of publishing intermediate datasets as citable products. {[}\protect\hyperlink{ref-HbZ13t8l}{140}/{]}
The pieces discuss the value of small curations done both in systematic reviews and by experimentalists in the course of their research projects. Published curation tables can serve as a source for improving the ecosystem of open knowledge, not less by reconciliation to Wikidata (thereby bridging the commentary with this project)

\hypertarget{wisecube---enterprise-biomedical-question-and-answering}{%
\subsubsection{WiseCube - enterprise biomedical question and answering}\label{wisecube---enterprise-biomedical-question-and-answering}}

During a part of this project, I have worked part-time as a consultant for the Wisecube company, based in Seattle, United States. {[}\protect\hyperlink{ref-gdYsBE7d}{141}/{]}
The job was approved by FAPESP, and consisted mainly in writing SPARQL queries that probe Wikidata for answers to the questions posed by the BioASQ competition. {[}\protect\hyperlink{ref-rkXotO9x}{142}{]}
It also entails on-demand curation of biomedical topics on Wikidata based on requests by pharmaceutical companies as well as the development of dashboards targeted at providing insights to customers.

\hypertarget{awards-and-participation-in-events}{%
\subsection{Awards and participation in events}\label{awards-and-participation-in-events}}

During the initial course of this PhD project, I have participated in several events:

\begin{itemize}
\tightlist
\item
  (Feb-2021) Presented an open talk at the ``Semana da Bioinformática'' event about modelling of biological systems (1020 views as of December 2021) {[}\protect\hyperlink{ref-1C9uHr1Zk}{143}, =VDvCxskIGEI{]}
\item
  (Jun-Aug 2021) Helped to organize the No-Budget-Science HackWeek virtual hackathon {[}\protect\hyperlink{ref-12LOzmXRs}{144}{]}
\item
  (Jul - 2021) Presented the work ``Wikidata for 5-star Linked Open Databases: A case study of PanglaoDB'' at the Bio-Ontologies section of the Annual International Conference on Intelligent Systems for Molecular Biology. {[}\protect\hyperlink{ref-SALI6Ywb}{145}{]}. The presentation was awarded the best
\item
  (Jul - 2021) Awarded the 2nd place in the International Society for Computational Biology (ISCB) Wikipedia Competition for the contributions to the Wikipedia page on Biocuration (\url{https://en.wikipedia.org/wiki/Biocuration}) {[}\protect\hyperlink{ref-IJG65hFm}{146}{]}
\item
  (Nov - 2021) Managed a project during BioHackathon Europe 2021, in Barcelona, Spain, on the representation of ELIXIR information on Wikidata. {[}\protect\hyperlink{ref-14Wi842eZ}{147}{]}
\end{itemize}

\hypertarget{course-work}{%
\subsection{Course work}\label{course-work}}

During the first year of the PhD program

@@

\leavevmode\vadjust pre{\hypertarget{ref-xWhvkd3B}{}}%
\CSLLeftMargin{22. }
\CSLRightInline{\textbf{An estimation of the number of cells in the human body}
\CSLBlock{Eva Bianconi, Allison Piovesan, Federica Facchin, Alina Beraudi, Raffaella Casadei, Flavia Frabetti, Lorenza Vitale, Maria Chiara Pelleri, Simone Tassani, Francesco Piva, \ldots{} Silvia Canaider} \emph{Annals of Human Biology} (2013-07-05) \url{https://www.wikidata.org/wiki/Q34037445}
\CSLBlock{DOI: \href{https://doi.org/10.3109/03014460.2013.807878}{10.3109/03014460.2013.807878}}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-MxIeSJYt}{}}%
\CSLLeftMargin{135. }
\CSLRightInline{\textbf{fcoex: FCBF-based Co-Expression Networks for Single Cells}
\CSLBlock{Tiago Lubiana, Helder Nakaya} \emph{Bioconductor version: Release (3.14)} (2021) \url{https://bioconductor.org/packages/fcoex/}}

\leavevmode\vadjust pre{\hypertarget{ref-CQRJ53gu}{}}%
\CSLLeftMargin{136. }
\CSLRightInline{\textbf{Complex Portal 2022: new curation frontiers}
\CSLBlock{Birgit HM Meldal, Livia Perfetto, Colin Combe, Tiago Lubiana, João Vitor Ferreira~Cavalcante, Hema Bye-A-Jee, Andra Waagmeester, Noemi del-Toro, Anjali Shrivastava, Elisabeth Barrera, \ldots{} Sandra Orchard} \emph{Nucleic Acids Research} (2021-10-29) \url{https://www.wikidata.org/wiki/Q109348309}
\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkab991}{10.1093/nar/gkab991}}}

\leavevmode\vadjust pre{\hypertarget{ref-1DguATd9G}{}}%
\CSLLeftMargin{137. }
\CSLRightInline{\textbf{The Cellosaurus, a cell-line knowledge resource.}
\CSLBlock{Amos Bairoch} \emph{Journal of Biomolecular Techniques} (2018-05-01) \url{https://www.wikidata.org/wiki/Q54370168}
\CSLBlock{DOI: \href{https://doi.org/10.7171/jbt.18-2902-002}{10.7171/jbt.18-2902-002}}}

\leavevmode\vadjust pre{\hypertarget{ref-lMQxhx3q}{}}%
\CSLLeftMargin{138. }
\CSLRightInline{\textbf{User:CellosaurusBot - Wikidata} \url{https://www.wikidata.org/wiki/User:CellosaurusBot}}

\leavevmode\vadjust pre{\hypertarget{ref-F2mYjDJ0}{}}%
\CSLLeftMargin{139. }
\CSLRightInline{\textbf{The Brazilian Reproducibility Initiative}
\CSLBlock{Ana P Wasilewska-Sampaio, Olavo Bohrer Amaral, Kleber Neves, Ana P Wasilewska-Sampaio, Clarissa FD Carneiro, Olavo Bohrer Amaral, Clarissa FD Carneiro} \emph{eLife} (2019-02-05) \url{https://www.wikidata.org/wiki/Q61799268}
\CSLBlock{DOI: \href{https://doi.org/10.7554/elife.41602}{10.7554/elife.41602}}}

\leavevmode\vadjust pre{\hypertarget{ref-HbZ13t8l}{}}%
\CSLLeftMargin{140. }
\CSLRightInline{\url{https://osf.io/preprints/metaarxiv/vbwa9}}

\leavevmode\vadjust pre{\hypertarget{ref-gdYsBE7d}{}}%
\CSLLeftMargin{141. }
\CSLRightInline{\textbf{Wisecube AI \textbar{} Knowledge Graph Engine} \url{https://www.wisecube.ai/}}

\leavevmode\vadjust pre{\hypertarget{ref-rkXotO9x}{}}%
\CSLLeftMargin{142. }
\CSLRightInline{\textbf{An overview of the BIOASQ large-scale biomedical semantic indexing and question answering competition}
\CSLBlock{George Tsatsaronis, Georgios Balikas, Prodromos Malakasiotis, Ioannis Partalas, Matthias Zschunke, Michael R Alvers, Dirk Weissenborn, Anastasia Krithara, Sergios Petridis, Dimitris Polychronopoulos, \ldots{} Georgios Paliouras} \emph{BMC Bioinformatics} (2015-04-30) \url{https://www.wikidata.org/wiki/Q28646342}
\CSLBlock{DOI: \href{https://doi.org/10.1186/s12859-015-0564-6}{10.1186/s12859-015-0564-6}}}

\leavevmode\vadjust pre{\hypertarget{ref-1C9uHr1Zk}{}}%
\CSLLeftMargin{143. }
\CSLRightInline{\textbf{YouTube} \url{https://www.youtube.com/}}

\leavevmode\vadjust pre{\hypertarget{ref-12LOzmXRs}{}}%
\CSLLeftMargin{144. }
\CSLRightInline{\textbf{No Budget Science Hack Week}
\CSLBlock{reprodutibilidade} \url{https://www.reprodutibilidade.bio.br/hack-week-2021}}

\leavevmode\vadjust pre{\hypertarget{ref-SALI6Ywb}{}}%
\CSLLeftMargin{145. }
\CSLRightInline{\textbf{Wikidata for 5-star Linked Open Databases: A case study of PanglaoDB}
\CSLBlock{Tiago Lubiana, João Vitor Ferreira Cavalcante} \emph{Zenodo} (2021-12-01) \url{https://doi.org/gnpzvr}
\CSLBlock{DOI: \href{https://doi.org/10.5281/zenodo.5747849}{10.5281/zenodo.5747849}}}

\leavevmode\vadjust pre{\hypertarget{ref-IJG65hFm}{}}%
\CSLLeftMargin{146. }
\CSLRightInline{\textbf{Biocuration - Wikipedia} \url{https://en.wikipedia.org/wiki/Biocuration}}

\leavevmode\vadjust pre{\hypertarget{ref-14Wi842eZ}{}}%
\CSLLeftMargin{147. }
\CSLRightInline{\textbf{biohackathon-projects-2021/projects/32 at main · elixir-europe/biohackathon-projects-2021}
\CSLBlock{GitHub} \url{https://github.com/elixir-europe/biohackathon-projects-2021}}
